---
title: "DATA 606: Final Project"
author: "Puja Roy"
date: "5/10/24"
output: openintro::lab_report
---

### Introduction

 I leveraged the Drug Use by Age dataset from FiveThirtyEight https://github.com/fivethirtyeight/data. The reason I chose this dataset is because many people are addicted to drugs ranging from multiple ages and I wanted to explore which ages and what types of drugs certain people are addicted to. 
 
### Abstract

This final project utilizes the "Drug Use by Age" dataset from FiveThirtyEight to explore the relationship between age and drug use. The primary research question investigates how drug use varies by age and identifies significant trends and patterns for specific substances. The dataset comprises survey data from the National Survey on Drug Use and Health, covering 17 distinct age groups ranging from 12 to 65+. The study is observational, with age as the independent variable and the percentage of users for various drugs as the dependent variables. 

Using linear regression models, I analyzed the use of substances such as crack, alcohol, and heroin across different age groups. Each model revealed a straight line of best fit, indicating linear relationships between age and drug use for these substances. The analysis provided insights into how the propensity for drug use changes with age, highlighting significant relationships where appropriate. 

By examining these patterns, the study aims to enhance understanding of drug addiction across age demographics, providing valuable information for public health strategies and interventions. The data and findings underscore the importance of age-specific approaches in addressing substance abuse.
 
```{r load-packages, message=FALSE}
#install.packages("DT")
library(DT)
library(tidyverse)
library(openintro)
library(dplyr)
library(tidyr)
library(psych)
library(ggplot2)
```

### Load the Data

```{r}
# Load the data
drug_data <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/drug-use-by-age/drug-use-by-age.csv")

# View the structure of the data
str(drug_data)

# Remove NA values
clean_drug_data <- na.omit(drug_data)
```

```{r}
# View the structure of the cleaned data
str(clean_drug_data)

# Summary statistics for the cleaned data
summary(clean_drug_data)

```

```{r}
# Summary statistics for quantitative variables
describe(clean_drug_data)
```

```{r}
# Summary statistics by age group
describeBy(clean_drug_data, group = drug_data$age)
```

```{r}
# Frequency table for age group
age_table <- table(clean_drug_data$age)
print(age_table)
```

```{r}
# Print column names
colnames(clean_drug_data)
```


### Exploratory Data Analysis (EDA)

```{r}
datatable(clean_drug_data)
```

### Data Visualizations

I used density plots to visualize the distribution of drug use percentages across different age groups. They provide a smooth, continuous representation of the data's distribution, allowing for easier identification of patterns and trends that may not be as apparent in histograms or other types of plots.

### Density Plots

```{r}
# Density plots for each drug
density_plots <- lapply(names(clean_drug_data)[3:ncol(clean_drug_data)], function(drug_name) {
  ggplot(clean_drug_data, aes_string(x = drug_name)) +
    geom_density(fill = "skyblue", color = "black") +
    labs(title = paste("Density Plot of", drug_name, "Use"),
         x = paste(drug_name, "Use"),
         y = "Density")
})

# Output the density plots
density_plots

```

### QQ Plots

I used QQ plots help assess the normality of the data distributions.

```{r}
# QQ plots for each drug
qq_plots <- lapply(names(clean_drug_data)[3:ncol(clean_drug_data)], function(drug_name) {
  ggplot(clean_drug_data, aes_string(sample = drug_name)) +
    stat_qq() +
    stat_qq_line() +
    labs(title = paste("QQ Plot of", drug_name, "Use"),
         x = "Theoretical Quantiles",
         y = paste("Sample Quantiles of", drug_name))
})

# Output the QQ plots
qq_plots

```


### Summary Statistics

The summary statistics for each drug and age group provides insights into the typical usage patterns, the range of usage, and the presence of any outliers.

```{r}
# Summary statistics for each drug
summary_stats <- lapply(names(clean_drug_data)[3:ncol(clean_drug_data)], function(drug_name) {
  summary(clean_drug_data[[drug_name]])
})

# Output the summary statistics
summary_stats

```

### Linear Regression Models

To analyze the relationship between age and drug use, I created linear regression models for various substances, including cocaine, crack, alcohol, and heroin. The linear regression model for each drug helps to understand how drug use percentage changes with age.

Null Hypothesis (H0): Age has no effect on the percentage of users of inhalant use.
Alternative Hypothesis (H1): Age has a significant effect on the percentage of users of Marijuana.

Cocaine Use:
H0: Age has no effect on the percentage of cocaine users.
H1: Age has a significant effect on the percentage of cocaine users.

```{r}
# Remove rows with any NA, NaN, or Inf values
clean_drug_data <- drug_data %>%
  mutate(across(everything(), ~ifelse(is.nan(.), NA, .))) %>%
  filter_all(all_vars(!is.na(.))) %>%
  filter_all(all_vars(!is.infinite(.)))

# Verify the cleaned data
summary(clean_drug_data)

```

```{r}
# Create a binary variable for marijuana use (e.g., use > 0%)
clean_drug_data$marijuana_use_binary <- ifelse(clean_drug_data$marijuana_use > 0, 1, 0)

# Check the distribution of the binary variable
table(clean_drug_data$marijuana_use_binary)

```

```{r}
# Fit a logistic regression model
logistic_model <- glm(marijuana_use_binary ~ age, data = clean_drug_data, family = binomial)

# Summary of the model
summary(logistic_model)

```

```{r}
# Convert age to numeric if necessary
drug_data$age <- as.numeric(as.character(drug_data$age))

# Remove rows with any NA, NaN, or Inf values
clean_drug_data <- drug_data %>%
  mutate(across(everything(), ~ifelse(is.nan(.), NA, .))) %>%
  filter_all(all_vars(!is.na(.))) %>%
  filter_all(all_vars(!is.infinite(.)))

# Verify the cleaned data
summary(clean_drug_data)

```

```{r}
# Create a binary variable for marijuana use (e.g., use > 0%)
clean_drug_data$marijuana_use_binary <- ifelse(clean_drug_data$marijuana_use > 0, 1, 0)

# Check the distribution of the binary variable
table(clean_drug_data$marijuana_use_binary)

```

```{r}
# Fit a logistic regression model
logistic_model <- glm(marijuana_use_binary ~ age, data = clean_drug_data, family = binomial)

# Summary of the model
summary(logistic_model)

```

```{r}
# Check for finite min and max values of age
min_age <- min(clean_drug_data$age, na.rm = TRUE)
max_age <- max(clean_drug_data$age, na.rm = TRUE)

# Generate data for prediction
age_seq <- seq(min_age, max_age, by = 1)
pred_data <- data.frame(age = age_seq)
pred_data$predicted_prob <- predict(logistic_model, newdata = pred_data, type = "response")

# Plot the logistic regression curve
ggplot(clean_drug_data, aes(x = age, y = marijuana_use_binary)) +
  geom_point(alpha = 0.5) +
  geom_line(data = pred_data, aes(x = age, y = predicted_prob), color = "red") +
  labs(title = "Logistic Regression: Probability of Marijuana Use by Age",
       x = "Age",
       y = "Probability of Marijuana Use")

```

```{r}
# Convert age to numeric if necessary
drug_data$age <- as.numeric(as.character(drug_data$age))

# Remove rows with any NA, NaN, or Inf values
clean_drug_data <- drug_data %>%
  mutate(across(everything(), ~ifelse(is.nan(.), NA, .))) %>%
  filter_all(all_vars(!is.na(.))) %>%
  filter_all(all_vars(!is.infinite(.)))

# Verify the cleaned data
summary(clean_drug_data)

```

```{r}
# Fit a linear regression model for cocaine use
linear_model_cocaine <- lm(cocaine_use ~ age, data = clean_drug_data)

# Summary of the model
summary(linear_model_cocaine)

```

```{r}
# Generate predictions for the plot
clean_drug_data$predicted_cocaine_use <- predict(linear_model_cocaine, newdata = clean_drug_data)

# Plot the linear regression model
ggplot(clean_drug_data, aes(x = age, y = cocaine_use)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = predicted_cocaine_use), color = "red") +
  labs(title = "Linear Regression: Cocaine Use by Age",
       x = "Age",
       y = "Cocaine Use (%)") +
  theme_minimal()

```

```{r}
# Fit a linear regression model for cocaine use
linear_model_alcohol <- lm(alcohol_use ~ age, data = clean_drug_data)

# Summary of the model
summary(linear_model_alcohol)

```

```{r}
# Generate predictions for the plot
clean_drug_data$predicted_alcohol_use <- predict(linear_model_alcohol, newdata = clean_drug_data)

# Plot the linear regression model
ggplot(clean_drug_data, aes(x = age, y = cocaine_use)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = predicted_alcohol_use), color = "red") +
  labs(title = "Linear Regression: Alcohol Use by Age",
       x = "Age",
       y = "Alcohol Use (%)") +
  theme_minimal()

```

```{r}
# Fit a linear regression model for cocaine use
linear_model_heroin <- lm(heroin_use ~ age, data = clean_drug_data)

# Summary of the model
summary(linear_model_heroin)

```

```{r}
# Generate predictions for the plot
clean_drug_data$predicted_heroin_use <- predict(linear_model_heroin, newdata = clean_drug_data)

# Plot the linear regression model
ggplot(clean_drug_data, aes(x = age, y = heroin_use)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = predicted_heroin_use), color = "red") +
  labs(title = "Linear Regression: Heroin Use by Age",
       x = "Age",
       y = "Heroin Use (%)") +
  theme_minimal()

```

```{r}
# Fit a linear regression model for cocaine use
linear_model_crack <- lm(crack_use ~ age, data = clean_drug_data)

# Summary of the model
summary(linear_model_crack)

```

```{r}
# Generate predictions for the plot
clean_drug_data$predicted_crack_use <- predict(linear_model_crack, newdata = clean_drug_data)

# Plot the linear regression model
ggplot(clean_drug_data, aes(x = age, y = crack_use)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = predicted_crack_use), color = "red") +
  labs(title = "Linear Regression: Crack Use by Age",
       x = "Age",
       y = "Crack Use (%)") +
  theme_minimal()

```

```{r}
# Fit linear regression models for each substance use
linear_model_crack <- lm(crack_use ~ age, data = clean_drug_data)
linear_model_alcohol <- lm(alcohol_use ~ age, data = clean_drug_data)
linear_model_heroin <- lm(heroin_use ~ age, data = clean_drug_data)
linear_model_marijuana <- lm(marijuana_use ~ age, data = clean_drug_data)
linear_model_sedative <- lm(sedative_use ~ age, data = clean_drug_data)
linear_model_painreliever <- lm(pain_releiver_use ~ age, data = clean_drug_data)
linear_model_inhalant <- lm(inhalant_use ~ age, data = clean_drug_data)

# Summarize each model
summary(linear_model_crack)
summary(linear_model_alcohol)
summary(linear_model_heroin)
summary(linear_model_marijuana)
summary(linear_model_sedative)
summary(linear_model_painreliever)
summary(linear_model_inhalant)

# Generate predictions for the plots
clean_drug_data$predicted_crack_use <- predict(linear_model_crack, newdata = clean_drug_data)
clean_drug_data$predicted_alcohol_use <- predict(linear_model_alcohol, newdata = clean_drug_data)
clean_drug_data$predicted_heroin_use <- predict(linear_model_heroin, newdata = clean_drug_data)
clean_drug_data$predicted_marijuana_use <- predict(linear_model_marijuana, newdata = clean_drug_data)
clean_drug_data$predicted_sedative_use <- predict(linear_model_sedative, newdata = clean_drug_data)
clean_drug_data$predicted_painreliever_use <- predict(linear_model_painreliever, newdata = clean_drug_data)
clean_drug_data$predicted_inhalant_use <- predict(linear_model_inhalant, newdata = clean_drug_data)

# Plot the linear regression models
plot_crack <- ggplot(clean_drug_data, aes(x = age, y = crack_use)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = predicted_crack_use), color = "red") +
  labs(title = "Linear Regression: Crack Use by Age",
       x = "Age",
       y = "Crack Use (%)") +
  theme_minimal()

plot_alcohol <- ggplot(clean_drug_data, aes(x = age, y = alcohol_use)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = predicted_alcohol_use), color = "red") +
  labs(title = "Linear Regression: Alcohol Use by Age",
       x = "Age",
       y = "Alcohol Use (%)") +
  theme_minimal()

plot_heroin <- ggplot(clean_drug_data, aes(x = age, y = heroin_use)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = predicted_heroin_use), color = "red") +
  labs(title = "Linear Regression: Heroin Use by Age",
       x = "Age",
       y = "Heroin Use (%)") +
  theme_minimal()

plot_marijuana <- ggplot(clean_drug_data, aes(x = age, y = marijuana_use)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = predicted_heroin_use), color = "red") +
  labs(title = "Linear Regression: Marijuana Use by Age",
       x = "Age",
       y = "Marijuana Use (%)") +
  theme_minimal()

plot_sedative <- ggplot(clean_drug_data, aes(x = age, y = sedative_use)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = predicted_sedative_use), color = "red") +
  labs(title = "Linear Regression: Sedative Use by Age",
       x = "Age",
       y = "Sedative Use (%)") +
  theme_minimal()

plot_painreliever <- ggplot(clean_drug_data, aes(x = age, y = pain_releiver_use)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = predicted_painreliever_use), color = "red") +
  labs(title = "Linear Regression: Pain Reliever Use by Age",
       x = "Age",
       y = "Pain Reliever Use (%)") +
  theme_minimal()

plot_inhalant <- ggplot(clean_drug_data, aes(x = age, y = inhalant_use)) +
  geom_point(alpha = 0.5) +
  geom_line(aes(y = predicted_inhalant_use), color = "red") +
  labs(title = "Linear Regression: Inhalant Use by Age",
       x = "Age",
       y = "Inhalant Use (%)") +
  theme_minimal()

# Output the plots
plot_crack
plot_alcohol
plot_heroin
plot_marijuana
plot_sedative
plot_painreliever
plot_inhalant

```

```{r}
# Summarize each model
summary_crack <- summary(linear_model_crack)
summary_alcohol <- summary(linear_model_alcohol)
summary_heroin <- summary(linear_model_heroin)
summary_marijuana <- summary(linear_model_marijuana)

# Extract the relevant statistics
# For crack use
t_value_crack <- summary_crack$coefficients["age", "t value"]
p_value_crack <- summary_crack$coefficients["age", "Pr(>|t|)"]
conf_int_crack <- confint(linear_model_crack)["age", ]

# For alcohol use
t_value_alcohol <- summary_alcohol$coefficients["age", "t value"]
p_value_alcohol <- summary_alcohol$coefficients["age", "Pr(>|t|)"]
conf_int_alcohol <- confint(linear_model_alcohol)["age", ]

# For heroin use
t_value_heroin <- summary_heroin$coefficients["age", "t value"]
p_value_heroin <- summary_heroin$coefficients["age", "Pr(>|t|)"]
conf_int_heroin <- confint(linear_model_heroin)["age", ]

# For marijuana use
t_value_marijuana <- summary_marijuana$coefficients["age", "t value"]
p_value_marijuana <- summary_marijuana$coefficients["age", "Pr(>|t|)"]
conf_int_marijuana <- confint(linear_model_marijuana)["age", ]

# Output the results
list(
  crack_use = list(
    t_value = t_value_crack,
    p_value = p_value_crack,
    conf_int = conf_int_crack
  ),
  alcohol_use = list(
    t_value = t_value_alcohol,
    p_value = p_value_alcohol,
    conf_int = conf_int_alcohol
  ),
  heroin_use = list(
    t_value = t_value_heroin,
    p_value = p_value_heroin,
    conf_int = conf_int_heroin
  ),
  marijuana_use = list(
    t_value = t_value_marijuana,
    p_value = p_value_marijuana,
    conf_int = conf_int_marijuana
  )
)
```

### Data Analysis

Based on the linear regression models, it is evident that there was a linear relationship between age and the use of crack, heroin, and alcohol. For each of these substances, the models demonstrated significant trends with age as a predictor.

Crack Use

The linear regression model for crack use revealed a positive relationship between age and the percentage of users. This suggests that as age increases, the percentage of individuals using crack also tends to increase. The statistical significance of this relationship indicates that age is an important factor in understanding crack use patterns.


### Conclusion

To analyze the relationship between age and drug use, I created linear regression models for various substances, including cocaine, crack, alcohol, and heroin. Each model aimed to understand how drug use percentages change with age. The summary output for each model provided key metrics such as coefficients, R-squared value, p-value, and confidence intervals. These metrics helped us evaluate the significance and strength of the relationship between age and drug use. These findings highlight the importance of age-specific interventions and policies. Younger age groups might benefit from different preventive measures compared to older groups. Public health officials and policymakers can use this information to design targeted strategies to reduce drug use among specific age demographics. Tailored education and prevention programs can be more effective in addressing drug abuse.



