---
title: "DATA 606 Data Project Proposal"
author: "Puja Roy"
---

### Data Preparation

For this final project proposal, I am planning to leverage the **Drug Use by Age** dataset from FiveThirtyEight https://github.com/fivethirtyeight/data. The reason I chose this dataset is because many people are addicted to drugs ranging from multiple ages and I wanted to explore which ages and what types of drugs certain people are addicted to.


```{r}
# Load necessary libraries
library(tidyverse)

# Load the data
drug_data <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/drug-use-by-age/drug-use-by-age.csv")

# View the structure of the data
str(drug_data)

# Check the first few rows of the data
head(drug_data)

```


### Research question 

**How does drug use vary by age, and are there any significant relationships, trends or patterns based on certain types of drugs?**


### Cases 

**What are the cases, and how many are there?**

The cases are different age groups, ranging from 12 to 65+. There are 17 cases in total.


### Data collection 

**Describe the method of data collection.**

The data was collected through surveys conducted by FiveThirtyEight.
Source: National Survey on Drug Use and Health from the Substance Abuse and Mental Health Data Archive.

### Type of study 

**What type of study is this (observational/experiment)?**

This study is observational.


### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

The data was collected by FiveThirtyEight and is publicly available on their GitHub repository.
Source: National Survey on Drug Use and Health from the Substance Abuse and Mental Health Data Archive.


### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

The response variable is the percentage of people who use each drug within different age groups. It is quantitative.

### Independent Variable(s)

Age is the independent variable in this study.


### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}
# Summary statistics
summary(drug_data)

# Visualizations
# Boxplot of marijuana use by age
ggplot(drug_data, aes(x = as.factor(age), y = marijuana_use)) +
  geom_point() +
  labs(title = "Marijuana Use by Age Group",
       x = "Age Group",
       y = "Percentage of Users")

# Scatterplot of cocaine use by age
ggplot(drug_data, aes(x = age, y = cocaine_use)) +
  geom_point() +
  labs(title = "Cocaine Use by Age",
       x = "Age",
       y = "Percentage of Users")

```

